Paper figures

### 2019-01-07 paper_figures.Rmd
Edit captions and img size

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, paged.print=FALSE, echo=FALSE)
knitr::opts_chunk$set(fig.width=12, fig.height=8) 
library(phyloseq)
library(magrittr)
library(tidyverse)
library(reshape2)
library(ggsignif)
library(DESeq2)

@@

# Figure 3
## Panel A: Alpha diversity controls T1 vs T2 and pre and post challenge T2 vs T3 
```{r Calculate alpha diversity, cache=TRUE}
alpha_div <- estimate_richness(gg_topi) %>% bind_cols(data.frame(sample_data(gg_topi)))
```

@@

```{r Plot boxplot}
filt_alpha %>%
  melt(id.vars = c('Description','condizione','timepoint'), measure.vars =  c('Observed', 'Chao1', 'Simpson', 'Shannon')) %>%
  ggplot() +
  geom_boxplot(aes(timepoint, value)) +
  xlab('Timepoint') +
  ylab('Value') +
  facet_grid(variable~.,scales = 'free_y') +
  theme_minimal() +
  theme( strip.background = element_rect(colour = "transparent", fill = "transparent"),
           strip.text.y = element_text(angle = 0))
```

## Panel B: Beta diversity differences pre and post challange T2 vs T3
```{r, cache=TRUE}
distwu <- phyloseq::distance(gg_topi, "wunifrac")
ordwu <-  ordinate(gg_topi, method = "MDS", distance = distwu, weighted = TRUE) %>% plot_ordination(gg_topi, ., justDF = TRUE)
distanceswu <- calculate_pairwise_distances(distwu)
distancesbc <- calculate_pairwise_distances(phyloseq::distance(gg_topi, "bray"))
metadata_2cat <- mappingfile[mappingfile$timepoint %in% c('T2','T3'),]
distwu_2cat <- phyloseq::distance(subset_samples(gg_topi, timepoint %in% c('T2','T3')), "wunifrac")
permanova <- vegan::adonis(formula = distwu_2cat ~ merge_timepoint, data = metadata_2cat)
```

```{r Plot ordination}
subset(ordwu, timepoint %in% c('T2','T3')) %>%
ggplot(aes(Axis.1, Axis.2)) +
  geom_point(aes(color=timepoint), size=3, alpha=.8) +
  scale_color_brewer(name = "Timepoint", palette="Dark2") + 
  theme_minimal() +
  annotation_custom(grob = grid::textGrob(label = bquote(atop("p-value"==.(permanova$aov.tab$`Pr(>F)`[1]),R^2==.(signif(permanova$aov.tab$R2[1], 2)))),
    gp=grid::gpar(col="black", fontsize=13)),  xmin = 0.15, xmax = .15, ymin = -.05, ymax = -.05)
```

## Panel C: Differential abundance analysis highlighted Bacteroidales S24/7 and Lacnospiraceae as most differential abunant families.
```{r DESeq2 loading, cache=TRUE}
pds_genus <- phyloseq_to_deseq2(gg_topi.genus, design = ~timepoint)
dds_genus <- DESeq2::DESeq(pds_genus, parallel = TRUE, quiet = TRUE)
```

@@

```{r}
  lacno_bact_relabs %>%
  ggplot() +
    geom_boxplot(aes(timepoint, Abundance, fill=timepoint)) + 
    facet_wrap(interaction(Family, Genus) ~ ., scales = "free_x", as.table= T) +
    theme_minimal() +
    guides(fill=FALSE) +
    theme( strip.background = element_rect(colour = "transparent", fill = "transparent"),
           strip.text.x = element_text(size=12))

@@

## Panel D: Ordination centroids beta diversity all timepoints 
```{r}
x <- ordinate(gg_topi, method = "MDS", distance = distwu, weighted = TRUE) %>% plot_ordination(gg_topi, ., justDF = TRUE)

@@

## Panel E: Boxplot of abundance of significant genera from DESeq2 differential analysis. Genera with median below 1% are filtered
```{r}
res37 <- DESeq2::results(dds_genus, name = "timepointT7", alpha = 0.05)
res37.filt <- res37[!is.na(res37$padj),]
res37.filt <- res37.filt[res37.filt$padj<0.05,]

@@

genus_ab_t %>%
  bind_cols(data.frame(tax_table(ggtopi_subs.genus.relab)[genus_ab_t$OTU,c('Phylum','Family','Genus')])) %>%
  filter(OTU %in% (genus_ab_t %>%
  group_by(OTU, timepoint) %>%
  summarise(median=median(Abundance)) %>%
  group_by(OTU) %>%
  summarise(filt=min(median)>0.01) %>% filter(filt==TRUE))$OTU) %>%
  filter(Phylum != "Unassigned") %>%
  mutate(taxlab = ifelse(stringr::str_detect(Genus,'uncultured'), paste('Uncultured',Family,sep = ' '), paste(Genus))) %>%
ggplot() +
  geom_boxplot(aes(reorder(taxlab, Abundance, FUN = median), Abundance, fill=timepoint)) +
  facet_grid(Phylum ~ ., scales = "free", space = "free_y", as.table= T) +
  scale_fill_brewer(name = "Timepoint", palette="Dark2") +
  coord_flip() +
  xlab('') +
  ylab("Relative abundance") +
  theme_cm()
```

## Panel F: Abundance of Bacteroidales 24/7 and Lachnospiraceae NK4A136 through timepoints and treatment
```{r}
idx_247 <- which(tax_table(gg_topi.genus.relab)[,'Family'] == "Bacteroidales S24-7 group" & tax_table(gg_topi.genus.relab)[,'Genus'] == "uncultured bacterium")
idx_lachno <- which(tax_table(gg_topi.genus.relab)[,'Genus'] == "Lachnospiraceae NK4A136 group" )

@@

## Panel G: Scatterplot of tumor size and difference of \\beta diversity 
```{r}
brayc_ds <- distancesbc %>% 
  dplyr::filter(topo1==topo2) %>%
  dplyr::filter(timepoint1==3) %>%
  dplyr::filter(timepoint2>4) %>%
  dplyr::select(cols, timepoint2, value) %>%
  dplyr::left_join(mappingfile, by=c('cols' = 'Description'))

@@

```{r corr tumor braycurtis}
distancesbc %>% 
  dplyr::filter(topo1==topo2) %>%
  dplyr::filter(timepoint1==3) %>%
  dplyr::filter(timepoint2>4) %>%
  dplyr::select(cols, timepoint2, value) %>%
  dplyr::left_join(mappingfile, by=c('cols' = 'Description')) %>%
  mutate(timepoint = paste('T',timepoint2,'-T3',sep = '')) %>%
  ggplot(aes(value,tumor_mass)) +
  geom_point() +
  facet_wrap(facets = 'timepoint', ncol = 1) +
  geom_line(stat='smooth', method = 'lm', color='red', alpha=0.7, size=.7) + 
  ylab(bquote('Tumor mass'~( mm^3))) +
  xlab('Bray Curtis distance') +
  theme_cm() +
  geom_text(data = ds_corr_braycf, aes(x=0.4,y=2500, label=label), parse = TRUE, inherit.aes = FALSE)

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, paged.print=FALSE, echo=FALSE)
knitr::opts_chunk$set(fig.width=8, fig.height=6) 
library(phyloseq)
library(magrittr)
library(tidyverse)
library(reshape2)
library(ggsignif)
library(DESeq2)

@@

# Figure 3
```{r Calculate alpha diversity, cache=TRUE}
alpha_div <- estimate_richness(gg_topi) %>% bind_cols(data.frame(sample_data(gg_topi)))
```

@@

```{r Plot boxplot, fig.cap='Fig 3A: Alpha diversity controls T1 vs T2 and pre and post challenge T2 vs T3'}
filt_alpha %>%
  melt(id.vars = c('Description','condizione','timepoint'), measure.vars =  c('Observed', 'Chao1', 'Simpson', 'Shannon')) %>%
  ggplot() +
  geom_boxplot(aes(timepoint, value)) +
  xlab('Timepoint') +
  ylab('Value') +
  facet_grid(variable~.,scales = 'free_y') +
  theme_minimal() +
  theme( strip.background = element_rect(colour = "transparent", fill = "transparent"),
           strip.text.y = element_text(angle = 0))
```

```{r, cache=TRUE}
distwu <- phyloseq::distance(gg_topi, "wunifrac")
ordwu <-  ordinate(gg_topi, method = "MDS", distance = distwu, weighted = TRUE) %>% plot_ordination(gg_topi, ., justDF = TRUE)
distanceswu <- calculate_pairwise_distances(distwu)
distancesbc <- calculate_pairwise_distances(phyloseq::distance(gg_topi, "bray"))
metadata_2cat <- mappingfile[mappingfile$timepoint %in% c('T2','T3'),]
distwu_2cat <- phyloseq::distance(subset_samples(gg_topi, timepoint %in% c('T2','T3')), "wunifrac")
permanova <- vegan::adonis(formula = distwu_2cat ~ merge_timepoint, data = metadata_2cat)
```

```{r Plot ordination, fig.cap="Fig 3B: Beta diversity differences pre and post challange T2 vs T3"}
subset(ordwu, timepoint %in% c('T2','T3')) %>%
ggplot(aes(Axis.1, Axis.2)) +
  geom_point(aes(color=timepoint), size=3, alpha=.8) +
  scale_color_brewer(name = "Timepoint", palette="Dark2") + 
  theme_cm() +
  annotation_custom(grob = grid::textGrob(label = bquote(atop("p-value"==.(permanova$aov.tab$`Pr(>F)`[1]),italic(R^2)==.(signif(permanova$aov.tab$R2[1], 2)))),
    gp=grid::gpar(col="black", fontsize=13)),  xmin = 0.15, xmax = .15, ymin = -.05, ymax = -.05)
```

```{r DESeq2 loading, cache=TRUE}
pds_genus <- phyloseq_to_deseq2(gg_topi.genus, design = ~timepoint)
dds_genus <- DESeq2::DESeq(pds_genus, parallel = TRUE, quiet = TRUE)
```

@@

```{r fig.cap="Fig 3C: Differential abundance analysis highlighted Bacteroidales S24/7 and Lacnospiraceae as most differential abunant families."}
  lacno_bact_relabs %>%
  ggplot() +
    geom_boxplot(aes(timepoint, Abundance, fill=timepoint)) + 
    facet_wrap(interaction(Family, Genus) ~ ., scales = "free_x", as.table= T) +
    theme_minimal() +
    guides(fill=FALSE) +
    theme( strip.background = element_rect(colour = "transparent", fill = "transparent"),
           strip.text.x = element_text(size=12))

@@

```{r fig.cap="Fig 3D: Ordination centroids beta diversity all timepoints"}
x <- ordinate(gg_topi, method = "MDS", distance = distwu, weighted = TRUE) %>% plot_ordination(gg_topi, ., justDF = TRUE)

@@

```{r}
res37 <- DESeq2::results(dds_genus, name = "timepointT7", alpha = 0.05)
res37.filt <- res37[!is.na(res37$padj),]
res37.filt <- res37.filt[res37.filt$padj<0.05,]

@@

```{r fig.cap="Fig 3E: Boxplot of abundance of significant genera from DESeq2 differential analysis. Genera with median below 1% are filtered"}
genus_ab_t %>%
  bind_cols(data.frame(tax_table(ggtopi_subs.genus.relab)[genus_ab_t$OTU,c('Phylum','Family','Genus')])) %>%
  filter(OTU %in% (genus_ab_t %>%
  group_by(OTU, timepoint) %>%
  summarise(median=median(Abundance)) %>%
  group_by(OTU) %>%
  summarise(filt=min(median)>0.01) %>% filter(filt==TRUE))$OTU) %>%
  filter(Phylum != "Unassigned") %>%
  mutate(taxlab = ifelse(stringr::str_detect(Genus,'uncultured'), paste('Uncultured',Family,sep = ' '), paste(Genus))) %>%
ggplot() +
  geom_boxplot(aes(reorder(taxlab, Abundance, FUN = median), Abundance, fill=timepoint)) +
  facet_grid(Phylum ~ ., scales = "free", space = "free_y", as.table= T) +
  scale_fill_brewer(name = "Timepoint", palette="Dark2") +
  coord_flip() +
  xlab('') +
  ylab("Relative abundance") +
  theme_cm()
```

```{r fig.cap="Fig 3 F: Abundance of Bacteroidales 24/7 and Lachnospiraceae NK4A136 through timepoints and treatment"}
idx_247 <- which(tax_table(gg_topi.genus.relab)[,'Family'] == "Bacteroidales S24-7 group" & tax_table(gg_topi.genus.relab)[,'Genus'] == "uncultured bacterium")
idx_lachno <- which(tax_table(gg_topi.genus.relab)[,'Genus'] == "Lachnospiraceae NK4A136 group" )

@@

```{r}
brayc_ds <- distancesbc %>% 
  dplyr::filter(topo1==topo2) %>%
  dplyr::filter(timepoint1==3) %>%
  dplyr::filter(timepoint2>4) %>%
  dplyr::select(cols, timepoint2, value) %>%
  dplyr::left_join(mappingfile, by=c('cols' = 'Description'))

@@

```{r corr tumor braycurtis, fig.cap="Fig 3 G: Scatterplot of tumor size and difference of beta diversity "}
distancesbc %>% 
  dplyr::filter(topo1==topo2) %>%
  dplyr::filter(timepoint1==3) %>%
  dplyr::filter(timepoint2>4) %>%
  dplyr::select(cols, timepoint2, value) %>%
  dplyr::left_join(mappingfile, by=c('cols' = 'Description')) %>%
  mutate(timepoint = paste('T',timepoint2,'-T3',sep = '')) %>%
  ggplot(aes(value,tumor_mass)) +
  geom_point() +
  facet_wrap(facets = 'timepoint', ncol = 1) +
  geom_line(stat='smooth', method = 'lm', color='red', alpha=0.7, size=.7) + 
  ylab(bquote('Tumor mass'~( mm^3))) +
  xlab('Bray Curtis distance') +
  theme_cm() +
  geom_text(data = ds_corr_braycf, aes(x=0.4,y=2500, label=label), parse = TRUE, inherit.aes = FALSE)
